---
title: "Class 6: R functions HW"
author: "Canbin Cai (PID: A18087473)"
format: pdf
toc: true
---

# Q.6 in the worksheet  
Can you improve this analysis code? 

> Original code: 

```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
```

```{r}
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
```

```{r}
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
```

```{r}
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

# Improve the code above ---> **My Solution**. 

> **Steps on improving the code in Q.6:**  

1. I created the `function()` named **protein** with the argument of x, which will be the three kinase proteins *4AKE*, *1AKE*, and *1E4Y*. 

2. By using the `read.pdb()` function, it will access the online PDB file of the *bio3d package* that is being installed and access the data for the argument of x, which will be the proteins *4AKE*, *1AKE*, and *1E4Y*. And I named this **s**.     

3. By using the `trim.pdb()` function, it will access the specific protein chain and atom type we want to focus on. In this case, I want to only focus on the alpha carbon atoms in chain A of each protein structure in **s**, which will be proteins *4AKE*, *1AKE*, and *1E4Y*. And I named this **s.chain**. 

4. Then, I want to access the data frame in each *atom type* and *b-factors* column. And I named this **s.b**. 

5. Lastly, I want all the results that I generated above to be on graph plots using the `plotb3()` function.    


```{r}
library(bio3d)
protein <- function(x){
  s <-read.pdb(x)
  s.chain  <- trim.pdb(s, chain="A", elety="CA") 
  s.b <- s.chain$atom$b
  plotb3(s.b, sse=s.chain, typ="l", ylab="Bfactor")
}
```


Plot for kinase **4AKE**: 
```{r}
protein("4AKE")
```


Plot for kinase **1AKE**: 
```{r}
protein("1AKE")
```


Plot for kinase **1E4Y**: 
```{r}
protein("1E4Y")
```



